%!TEX root = ../../main.tex

\section{As Soon As Possible (ASAP) semantics}
In this section we introduce another semantics which captures the behaviour of systems that must response promptly.
This semantics is called As Soon As Possible (ASAP) semantics and forces a formula to satisfy a state-sequence in as few steps as possible.
To define the ASAP semantics we exploit another semantics, called \textit{bounded-steps semantics}, which forces a formula to be satisfied in a certain bound of time.
Similarly to the bounded-value semantics, we define the bounded-steps semantics for both co-safety and safety fragments of $\ltl$. 

In the co-safety fragment we say that a formula is satisfied by a state-sequence at position $i$ with maximum number of steps $u$, if such formula is satisfied in the prefix from $i$ to $u+i$.

\begin{definition}[Co-safety LTL bounded-steps semantics]
Let $\phi$ be a formula belonging to the co-safety fragment, $\sigma$ a state-sequence evaluated at position $i$ and $u \in \Nat$ maximum number of steps.
We say that $\sigma$ satisfies $\phi$ at position $i$ in $u$ steps, denoted as $\sigma,i \modelsUB{u}{} \phi$, if and only if
\begin{flalign*}
    \sigma[i,u+i] \models \phi
\end{flalign*}
We say that $\sigma \modelsUB{u}{} \phi$ if and only if $\sigma,0 \modelsUB{u}{} \phi$.
\end{definition}

In the safety fragment the definition is a little bit different.
First of all, we name the number of steps already performed as depth.
The difference between the upper bound and the depth gives us the number of steps that we are still allowed to do.
We need to understand what means for a state-sequence satisfying a formula in $u$ steps.
Here there are two possible interpretations: either defining such semantics similar to the semantics induced by the full-bounded dialect of $\ltl$ ($\fbltl$), therefore evaluating a property over a prefix from $i$ to $u+i$ like in the co-safety case, or evaluating over a finite prefix only formulas which are both in co-safety and safety.
In the first case we would define the semantics of formulas such as $\ltlGbounded{\phi}{[d,u]}$, $\ltlFbounded{\phi}{[d,u]}$ and $\ltlUbounded{\phi_1}{\phi_2}{[d,u]}$, but as we have already said, it is the same semantics induced by the full-bounded dialect of $\ltl$. 
Instead, we have preferred to investigate the latter case, where the semantics evaluates only both safety co-safety formulas over a prefix from $d$ to $u$. 
Analogously to the bounded-value semantics, the bounded-steps semantics is defined inductively on the structure of the formula to evaluate.
Basic cases are \textit{boolean propositions}, \textit{negation} of a boolean proposition, \textit{or} and \textit{and} operators, while more interesting cases are the \textit{next} and \textit{yesterday} temporal operators.

\begin{definition}[Safety LTL bounded-steps semantics]
Let $\phi$ be a formula belonging to the safety fragment, $\sigma$ a state-sequence evaluated at position $i$, $u \in \Nat$ maximum number of steps and $d \in \Nat$ be the depth such that $d \leq u$.
We say that $\sigma$ satisfies $\phi$ at position $i$ in $u$ steps with depth $d$, denoted as $\sigma,i \modelsUB{u}{d} \phi$, if and only if
\begin{flalign*}
&\sigma,i \modelsUB{u}{d} p &\iff& & & p \in \sigma_i \land 0 \leq d \leq u \\
&\sigma,i \modelsUB{u}{d} \ltlNeg{p} &\iff& & & p \not\in \sigma_i \land 0 \leq d \leq u \\
&\sigma,i \modelsUB{u}{d} \ltlOr{\phi_1}{\phi_2} &\iff& & & \text{$\sigma,i \modelsUB{u}{d} \phi_1$ or $\sigma,i \modelsUB{u}{d} \phi_2$} \\
&\sigma,i \modelsUB{u}{d} \ltlAnd{\phi_1}{\phi_2} &\iff& & & \text{$\sigma,i \modelsUB{u}{d} \phi_1$ and $\sigma,i \modelsUB{u}{d} \phi_2$} \\
&\sigma,i \modelsUB{u}{d} \ltlX{\phi} &\iff& & & \sigma,i+1 \modelsUB{u}{d+1} \phi \\
&\sigma,i \modelsUB{u}{d} \ltlY{\phi} &\iff& & & \text{$i>0$ and $\sigma,i-1 \modelsUB{u}{d-1} \phi$} \\
&\sigma,i \modelsUB{u}{d} \ltlR{\phi_1}{\phi_2} &\iff& & & \text{either for all $j \geq i. \; \sigma,j \modelsUB{u}{d} \phi_2$,} \\
& & & & & \text{or there exists $j \geq i$ such that $\sigma,j \modelsUB{u}{d} \phi_1$ and} \\
& & & & & \text{$\sigma,w \modelsUB{u}{d} \phi_2$ for all $i \leq w \leq j$} \\
&\sigma,i \modelsUB{u}{d} \ltlS{\phi_1}{\phi_2} & \iff & & & \text{there exists $0 \leq j \leq i$ such that $\sigma,j \modelsUB{u}{d} \phi_2$ and} \\
& & & & & \text{$\sigma,w \modelsUB{u}{d} \phi_1$ for all $j < w \leq i$} \\
&\sigma,i \modelsUB{u}{d} \ltlT{\phi_1}{\phi_2} & \iff & & & \text{either for all $0 \leq j \leq i. \; \sigma,j \modelsUB{u}{d} \phi_2$,} \\
& & & & & \text{or there exists $j \in [0,i]$ such that $\sigma,j \modelsUB{u}{d} \phi_1$ and} \\
& & & & & \text{$\sigma,w \modelsUB{u}{d} \phi_2$ for all $j \leq w \leq i$}
\end{flalign*}
We say that $\sigma,i \modelsUB{u}{} \phi$ if and only if $\sigma,i \modelsUB{u}{0} \phi$, while $\sigma \modelsUB{u}{} \phi$ if and only if $\sigma, 0 \modelsUB{u}{} \phi$.
\end{definition}

After having defined the bounded-steps semantics, we can finally present the \textit{ASAP semantics}, thanks to which we capture the fact to satisfy a formula in as few number of steps as possible.
The ASAP semantics is defined exploiting bounded-steps semantics and we require that the upper bound is as tight as possible, therefore it is the least upper bound of the bounds satisfying bounded-steps a formula.

\begin{definition}[ASAP semantics]
Let $\phi$ be a formula belonging to either the safety or co-safety fragment, $\sigma$ a state-sequence at position $i$ and $u \in \Nat$ maximum number of steps.
We say that $\sigma$ satisfies $\phi$ at position $i$ as soon as possible in $u$ steps, denoted as $\sigma, i \modelsASAP{u}{} \phi$, if and only if
\begin{flalign*}
    \sigma, i \modelsUB{u}{} \phi \land 
    \sigma, i \not\modelsUB{u-1}{} \phi
\end{flalign*}
We say that $\sigma \modelsASAP{u}{} \phi$ if and only if $\sigma, 0 \modelsASAP{u}{} \phi$.
\end{definition}

\input{chapters/original-contributions/asap-examples}
\input{chapters/original-contributions/asap-best-effort-same-semantics}
