%!TEX root = ../main.tex

\chapter{Original contributions} \label{chapt:original-contributions}

In this chapter we present all original contributions we made.

We formulate the problem to face, first textually and then formally. To do that we present four semantics for both safety and co-safety fragments: bounded-value, best-effort, bounded-steps and As Soon As Possible (ASAP).
Bounded-value semantics is the basis for best-effort one, while bounded-steps semantics for the ASAP one.
We prove that the ASAP semantics is just one of the possible instantiations of the best-effort semantics, showing that is possible to simulate the behaviour of ASAP semantics in the best-effort semantics.

The problem we want to deal with is an optimality test on a model, which we consider a closed-loop system for convenience.
In particular, given a plant, a controller and a safety or co-safety $\ltl$ formula, we want to figure out whether the closed-loop between plant and controller satisfies the formula and whether there exists another controller which does better than the given one according to some optimality principle. 
The optimality principle chosen by us is the one imposed by best-effort semantics because it is the most general one.
Then, we present a way to solve the problem by reduction to a reactive synthesis problem.

Finally, we provide a way to generate a non-deterministic strategy for reachability games.

\input{chapters/original-contributions/best-effort-semantics}
\input{chapters/original-contributions/asap-semantics}
\input{chapters/original-contributions/problem-formulation}
\input{chapters/original-contributions/reduction-to-synthesis}
\input{chapters/original-contributions/reachbility-strategy-extraction}
